---
title: "Working with HDFS"
author: "Hong Ooi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Working with HDFS}
  \usepackage[utf8]{inputenc}
---

<!-- to build this vignette as part of installation:
 - copy it to the /vignettes directory
 - ensure you are connected to a Spark or Hadoop cluster (a remote connection works)
 - run devtools::install(build_vignettes=TRUE) -->
 
```{r, echo=FALSE, message=FALSE, results="hide"}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
options(dplyr.print_min=5L, dplyr.print_max=5L)

dplyrXdf:::detectHdfsConnection()
```
Microsoft R Server includes the ability to work with Hadoop and Spark clusters. As part of this, it also lets you work with datasets, in particular Xdf files, that are stored in HDFS.

As of version 0.10.0, dplyrXdf also supports Xdf files in HDFS. Most verbs will work exactly the same as  with data in the native filesystem, and there are also a number of utility functions to ease working with Hadoop and Spark. You will need to have Microsoft R Server version 9.0 to use dplyrXdf with HDFS, and ideally version 9.1 for full functionality.


## Basics: Hadoop, Spark and HDFS

If you're used to dplyr and how it handles databases, Microsoft R Server and dplyrXdf work a little differently. What dplyr calls a "src" (a remote data source) is handled by two related concepts in MRS: a _compute context_ and a _filesystem_. The compute context determines where computations take place: on a local machine, in a database like SQL Server, or in a Hadoop or Spark cluster. The _filesystem_ is where the data files are stored. Note that not all data sources have a filesystem: eg a SQL Server table is part of a database, not a filesystem as such.

In general, there are only two filesystems you should have to deal with: the native filesystem, meaning the hard disk of the machine where your R session is running, and HDFS. To MRS and dplyrXdf, an Xdf file is much the same regardless of which filesystem it's in (there are some limitations for HDFS files, but they are the exception, not the rule). You can use the same code and pipelines without having to worry about filesystem details.

The compute context is important because it's how you benefit from the parallelism provided by Hadoop and Spark. Note that it doesn't change the way you _write_ your code -- only how it's executed. 

- If you are in the `RxHadoopMR` or `RxSpark` compute context, _and_ your data is in HDFS, then the RevoScaleR functions will execute in parallel on the cluster worker nodes. You can do this in an R session either on a remote client, or on the cluster edge node.

- If you are in the local compute context, and your R session is running on the cluster edge node, you can still access data in HDFS. In this case, the data is streamed to the edge node for processing -- essentially, RevoScaleR will treat HDFS as simply a big local hard disk. (This will not work if you are on a remote client; in this case, you must be in the Hadoop or Spark compute context to access the cluster.)

- If you are in the `RxHadoopMR` or `RxSpark` compute context and your data is _not_ in HDFS, then the RevoScaleR functions will (usually) throw a warning or error.

For more information about working with HDFS, Hadoop and Spark, see the [documentation at Microsoft Docs](https://docs.microsoft.com/en-us/r-server/r/how-to-revoscaler-spark), or the Microsoft Learn Analytics [course materials on Spark](https://github.com/Azure/LearnAnalytics-mrs-spark).


## Working with a cluster

To use dplyrXdf on a cluster, you'll have to install it, and dplyr, on the machine where you'll be running your R sessions. This can be either the cluster edge node, or if you're connecting remotely, on your remote machine. For full functionality, you should also install dplyr on the cluster worker nodes. You don't need to install dplyrXdf on the worker nodes, though.

Your dplyrXdf code should work the same regardless of whether you're running on the edge node or a remote client. The package abstracts away the underlying differences between the two scenarios, so that from your point of view, you're simply connected to a cluster somewhere in the cloud. Similarly, dplyrXdf should work with all flavours of Spark and Hadoop that MRS supports (HDInsight, Cloudera, Hortonworks and MapR).


### Uploading and downloading

There are a number of ways to get your data into and out of HDFS. Note that there are several data formats in use in Hadoop and Spark environments; dplyrXdf focuses mainly on Xdf files.

To copy a dataset from the native filesystem into HDFS, use `copy_to`. This is the standard dplyr verb for copying data into a remote src. In this case, the src is a filesystem, which RevoScaleR represents via a `RxHdfsFileSystem` object:

```{r}
library(dplyrXdf)
library(nycflights13)

hd <- RxHdfsFileSystem()

# copy a data frame into an Xdf file in HDFS
flightsHd <- copy_to(hd, flights)

flightsHd

as_data_frame(flightsHd)
```

This will copy the flights table into HDFS, returning an Xdf data source object pointing to the file. If the path isn't specified, the data is saved as an Xdf file in your HDFS home directory, normally `/user/<username>`.

You can also use `copy_to_hdfs`, which is a shortcut that saves having to create an explicit filesystem object:

```{r, eval=FALSE}
# same as above
flightsHd <- copy_to_hdfs(flights)
```

To download an Xdf file from HDFS to the native filesystem, use `collect` and `compute`. Again, these are the standard dplyr verbs for copying data from a remote src. For dplyrXdf, the difference between `collect` and `compute` is that the former by default will return a data frame, whereas the latter will save the copied data in an Xdf file and return a `tbl_xdf` object pointing to that file.

```{r}
flightsLocal <- compute(flightsHd)

flightsLocal

as_data_frame(flightsLocal)
```

If you want to look at the first few rows of an Xdf file, it may be faster to use `compute` to copy the entire file off HDFS, and then run `head`, than to run `head` on the original. This is due to quirks in how RevoScaleR works in Spark and Hadoop.

`copy_to`, `collect` and `compute` are for copying _datasets_ (R objects, including data frames and Xdf files) to and from HDFS. To transfer arbitrary files and directories, dplyrXdf also provides the `hdfs_upload` and `hdfs_download` functions. If you've used the base R function `download.file` or a command-line file transfer utility like `ftp` or `pscp`, the syntax should be familiar: `hdfs_download(src, dest)` downloads the file at `src` to the location `dest`, while `hdfs_upload(src, dest)` uploads `src` to `dest`.


```{r}
# create a csv file and upload it
write.csv(flights, "flights.csv", row.names=FALSE)
hdfs_upload("flights.csv", "/tmp")
```

If you uploaded a non-Xdf data source (like a csv file), you can then import it into Xdf format with `as_xdf`:

```{r}
flightsCsv <- RxTextData("/tmp/flights.csv", fileSystem=RxHdfsFileSystem())
flightsHd2 <- as_xdf(flightsCsv, "flights2")

as_data_frame(flightsHd2)
```

### Note on composite datasets

Note that there are two kinds of Xdf files: standard and _composite_. A composite Xdf file is actually a directory containing multiple data and metadata files. The RevoScaleR functions can treat a composite Xdf as a single dataset, and so can dplyrXdf. Xdf files in HDFS must be composite in order to work properly; by default, `copy_to` will convert an existing Xdf file into composite, if it's not already in that format.


## Sparklyr interoperability and Hive tables

If your R session is on the edge node, RevoScaleR provides the ability to share a Spark connection with the [sparklyr package](https://spark.rstudio.com) from RStudio. You do this via the `interop` argument to `rxSparkConnect`:

```{r, eval=FALSE}
# establishes a Spark connection that can be shared with sparklyr
rxSparkConnect(..., interop="sparklyr")
```

dplyrXdf can take advantage of this interoperability when working with Hive tables. To see how this works, let's use the `hivesampledata` table which is provided as part of every Azure HDInsight cluster:

<!-- to render this programmatically, you must be on the edge node -->

```{r, eval=FALSE}
cc <- rxSparkConnect(interop="sparklyr")

sampleHiv <- RxHiveData(table="hivesampletable")
head(sampleHiv, 5)
#>   clientid querytime market deviceplatform devicemake devicemodel        state
#> 1        8  18:54:20  en-US        Android    Samsung    SCH-i500   California
#> 2       23  19:19:44  en-US        Android        HTC  Incredible Pennsylvania
#> 3       23  19:19:46  en-US        Android        HTC  Incredible Pennsylvania
#> 4       23  19:19:47  en-US        Android        HTC  Incredible Pennsylvania
#> 5       28  01:37:50  en-US        Android   Motorola     Droid X     Colorado
#>         country querydwelltime sessionid sessionpagevieworder
#> 1 United States      13.920401         0                    0
#> 2 United States             NA         0                    0
#> 3 United States       1.475742         0                    1
#> 4 United States       0.245968         0                    2
#> 5 United States      20.309534         1                    1

sampleHiv %>%
    filter(deviceplatform == "Android") %>%
    group_by(devicemake) %>%
    summarise(n=n()) %>%
    arrange(desc(n)) %>%
    head()
#> # Source:     lazy query [?? x 2]
#> # Database:   spark_connection
#> # Ordered by: desc(n)
#>     devicemake     n
#>          <chr> <dbl>
#> 1      Samsung 16244
#> 2           LG  7950
#> 3          HTC  2242
#> 4      Unknown  2133
#> 5     Motorola  1524
#> # ... with more rows
```

In this example, dplyrXdf has converted the `RxHiveData` source into a sparklyr tbl, and the subsequent pipeline will be run by sparklyr (_not_ by dplyrXdf) in the cluster.

If you don't want to use sparklyr -- for example, if you want to take advantage of MRS features like a `transformFunc` to execute arbitrary R code -- you can import the Hive table to an Xdf file in HDFS. The pipeline will then execute according to your compute context. This will probably still be in-cluster, since you must be in the Spark compute context to access Hive tables. However, executing the pipeline via sparklyr is likely to be more efficient, since it removes the need to import the data.

```{r, eval=FALSE}
# this will create the composite Xdf 'samplehivetable' in the HDFS user directory
sampleXdf <- as_xdf(sampleHiv)

sampleXdf %>%
    filter(deviceplatform == "Android") %>%
    group_by(devicemake) %>%
    summarise(n=n()) %>%
    arrange(desc(n)) %>%
    head()
#>     devicemake     n
#> 1      Samsung 16244
#> 2           LG  7950
#> 3          HTC  2242
#> 4      Unknown  2133
#> 5     Motorola  1524
```


## Working with files and directories

dplyrXdf provides the following functions to let you manipulate files and directories in HDFS. By and large they wrap similar functions provided by the RevoScaleR package, which in turn call various Hadoop filesystem commands.

- `hdfs_file_copy(src, dest)` copies the file or directory given by `src` to the location `dest`. It is vectorised in both `src` and `dest`, meaning `src1` will be copied to `dest1`, `src2` to `dest2`, and so on. It is analogous to base R's `file.copy` for the native filesystem.
- `hdfs_file_move(src, dest)` is similar, but moves files/directories. It is analogous to base R's `file.rename`.
- `hdfs_file_remove` deletes the path. Its counterpart in base R is `file.remove`.
- `hdfs_dir_create` and `hdfs_dir_remove` create and delete directories. They are analogous to `dir.create` and `unlink(recursive=TRUE)`.
- `hdfs_file_exists` and `hdfs_dir_exists` test for the existence of a file or directory, like base R's `file.exists` and `dir.exists`.
- `hdfs_dir` lists files in a HDFS directory, returning a vector of file names. It has a number of options for recursively listing subdirectories, returning subdirectories only (omitting files), etc.

```{r}
# create a new directory
hdfs_dir_create("/tmp/mydata")

# check that it exists
hdfs_dir_exists("/tmp/mydata")

# copy files into the new directory
hdfs_file_copy("flights", "/tmp/mydata")

# create a new data source
flightsHd3 <- RxXdfData("/tmp/mydata/flights", fileSystem=RxHdfsFileSystem())

# read the data
names(flightsHd3)
```

### Azure Data Lake Store support

As with RevoScaleR, dplyrXdf supports HDFS data located in attached ADLS storage. When working with Xdf files, you will normally specify the HDFS filesystem as part of the `fileSystem` argument. The standard dplyr/dplyrXdf verbs can both read and write to ADLS, with the location of the file being essentially transparent to the user.

```{r, eval=FALSE}
adlFs <- RxHdfsFileSystem(hostName="adl://mycluster.azuredatalakestore.net")
adlXdf <- RxXdfData("/path/to/file", fileSystem=adlFs)

# adlXdf %>% verb1 %>% verb2 %>% ...
```

The `hdfs_*` functions include a `host` argument to let you specify which HDFS filesystem to access: either the "native" filesystem, or ADLS. The default will be whichever filesystem is currently being used. For example the following code uses `hdfs_dir` to list the files in the root directory, first for the native HDFS filesystem, and then for an attached ADLS filesystem.

<!-- to render this programmatically, your cluster must have ADLS storage attached -->

```{r, eval=FALSE}
hdfs_dir("/", host="default")
#> Directory listing of /
#>  [1] "HdiNotebooks"             "HdiSamples"
#>  [3] "ams"                      "amshbase"
#>  [5] "app-logs"                 "apps"
#>  [7] "atshistory"               "cluster-info"
#>  [9] "custom-scriptaction-logs" "example"
#> [11] "hbase"                    "hdp"
#> [13] "hive"                     "mapred"
#> [15] "mr-history"               "tmp"
#> [17] "user"

hdfs_dir("/", host="adl://mycluster.azuredatalakestore.net")
#> Directory listing of adl://mycluster.azuredatalakestore.net/
#> [1] "clusters"         "file31bc3ee52d32" "file31bc79a3de7"  "file5d301ef514e5"
#> [5] "folder1"          "tmp"              "user"
```

You can also provide a full URI as a path:

```{r, eval=FALSE}
hdfs_dir("adl://mycluster.azuredatalakestore.net/")
#> Directory listing of adl://mycluster.azuredatalakestore.net/
#> [1] "clusters"         "file31bc3ee52d32" "file31bc79a3de7"  "file5d301ef514e5"
#> [5] "folder1"          "tmp"              "user"
```

The `hdfs_host` function returns the name of the currently active HDFS filesystem host, or if an Xdf data source is supplied, the host for that data source:

```{r, eval=FALSE}
hdfs_host()
#> [1] "adl://mycluster.azuredatalakestore.net"

xdf <- RxXdfData("file", fileSystem=RxHdfsFileSystem(hostName="default"))
hdfs_host(xdf)
#> [1] "default"
```


## Miscellaneous functions

The `in_hdfs` function returns whether a given data source is stored in HDFS.

```{r}
in_hdfs(flightsHd)

# also works with non-Revo data sources, like data frames
in_hdfs(iris)

in_hdfs(flights)
```

The `local_exec` function runs an expression in the local compute context. This can be useful if you need to work with local datasets while connected to a remote cluster. By default, RevoScaleR functions will throw an error if you provide a local data source as an input when you are in the Hadoop or Spark compute context. `local_exec` temporarily changes to the local compute context, runs your code, and then changes back to the original context.

```{r, error=TRUE}
# try to access a local Xdf file
names(flightsLocal)

local_exec(names(flightsLocal))
```


## Limitations and glitches

Most of the single-table dplyr verbs supported by dplyrXdf will work for datasets in HDFS. The main exceptions are:

- `arrange`, `distinct`, `sample_n` and `sample_frac` do not support HDFS data
- `cbind` and `rbind` require the compute context to be local
- grouped `mutate` and `transmute` also require that the compute context is local. Consider whether you really need to group before transforming; many transformations do not require grouping information. If your data fits into memory, you can also use `do` or `do_xdf`.
- `summarise` on HDFS data will always send the output to the edge node/remote client, before writing it back to the cluster. This is a consequence of the way in which `rxCube` and `rxSummary` work, by creating an in-memory data structure.

Support for two-table verbs is similarly more limited for datasets in HDFS than in the native filesystem. First, only the Spark compute context supports joining (not Hadoop), and only for Xdf data sources and Spark data sources (`RxHiveData`, `RxOrcData` and `RxParquetData`). Only the "standard" joins (`left_join`, `right_join`, `inner_join` and `full_join`) are supported.

The RevoScaleR functions do not recognise `tbl_xdf` objects, which are automatically generated by dplyrXdf, when in the Spark or Hadoop compute context. To remedy this, wrap any such object inside `as_xdf()` before calling a Revo function:

```{r, eval=FALSE}
rxLinMod(y ~ x1 + x2 + ..., data=as_xdf(tbl))
```

Finally, the Xdf data sources created by dplyrXdf will always contain absolute HDFS paths (that is, of the form `"/path/to/file"` rather than `"file"`. This is because of limitations in RevoScaleR support for relative paths in HDFS. The data source _inputs_ to dplyrXdf verbs can contain relative paths; these will be internally translated to absolute.

```{r, echo=FALSE, message=FALSE, results="hide"}
hdfs_dir_remove(c("flights", "flights2", "/tmp/mydata", "/tmp/flights.csv"))
clean_dplyrxdf_dir("hdfs")
clean_dplyrxdf_dir("native")
```



